---
author: Emilio Berti
output: github_document
---

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.10890031.svg)](https://doi.org/10.5281/zenodo.10890031)

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 4,
  fig.height = 4,
  fig.align='center',
  dpi = 300,
  out.width = "50%"
)
```

# General information

The R package _parthian_ is developed and maintained by Emilio Berti (<emilio.berti@idiv.de>).
The scope of _parthian_ is to quantify the importance of areas in the landscape based on energy cost of movement for animals.
There are several dependencies for _parthian_:

  - Rcpp
  - igraph
  - terra
  - enerscape
  
They are all stable packages with long history, except for _enerscape_, which I developed in 2022 and maintain since then: <https://cran.r-project.org/package=enerscape> and <https://doi.org/10.1111/2041-210X.13734>.

```{r}
library(terra)
library(enerscape)
library(igraph)
library(parthian)
```

# Introduction

There are two datasets in _parthian_:

  - dem: a digital elevation model for an area in Sicily, Italy.
  - pa: the protected areas in the same region.

These are matrices, as it is easier to store them in an R package.
The first thing is to transform them into raster.

```{r torasters}
data(dem)
dem <- rast(
  dem,
  crs = "+proj=utm +zone=32 +datum=WGS84 +units=m +no_defs"
)
data(pa)
pa <- rast(
  pa,
  crs = "+proj=utm +zone=32 +datum=WGS84 +units=m +no_defs"
)
plot(dem, col = colorRampPalette(c("darkblue", "seagreen", "white"))(100))
plot(pa, add = TRUE, col = adjustcolor("gold", alpha.f = .5), legend = FALSE)
lines(as.polygons(pa))
```

The resolution and extent of the layers are wrong (I need to fix this in the package data), but it does not matter too much for the examples.

The next step is to calcualte the energy landscape for the animal.
Here, I am assuming an animal of 10 kg.

```{r enerscape}
en <- enerscape(dem, 10, "kcal")
plot(en, col = colorRampPalette(c("grey95", "tomato", "darkred"))(100))
plot(pa, add = TRUE, col = adjustcolor("gold", alpha.f = .5), legend = FALSE)
lines(as.polygons(pa))
```

The main task of _parthian_ is to create a graph where vertices (_V_) are the cells of the energy landscapes and weighted edges (_E_) $E_{ij} = E_{COT}$ if two cells are adjacent, and $E_{ij} = 0$ if they are not.

```{r cost-graph}
g <- cost_graph(en)
```

Generally, there are as many vertices as number of cells

```{r vertices}
length(V(g)) == ncell(en)
```

but the number of edges may be lower than $8V$, as some paths may be blocked, in this example by the sea.

```{r edges}
length(E(g)) == ncell(en) * 8
```

# Least cost paths
Least-cost paths can be obtained using the weighted graph created by `cost_graph()` and the _igraph_ `shortest_paths()` function.
First, let's get the centroids of the protected area, after exlcuding very small areas ($\leq 100 m^2$):

```{r centroids}
pas <- disagg(as.polygons(pa))
pas <- pas[expanse(pas, "m") > 100, ]
centrs <- centroids(pas)
plot(pa, col = "gold")
points(centrs, cex = 1, pch = 21)
points(centrs[c(1, 4), ], cex = 1, pch = 20)
```
Let's calcualte the least-cost path between the two areas highlighted by the solid circle.
Because there is a one-to-one correspondence between cell and vertex ID, this can be achieved by:

```{r lcp}
xy <- extract(en, centrs[c(1, 4), ], cells = TRUE)
lcp <- shortest_paths(g, xy$cell[1], xy$cell[2])
path <- lcp$vpath[[1]]
path <- xyFromCell(en, as.numeric(path))
path <- vect(path, crs = crs(dem))
total_costs <- sum(extract(en, path)[["EnergyScape"]])
```

```{r plot-path}
plot(en, col = colorRampPalette(c("grey95", "tomato", "darkred"))(100))
plot(pa, add = TRUE, col = adjustcolor("gold", alpha.f = .5), legend = FALSE)
lines(as.polygons(pa))
lines(as.lines(path), lw = 3, col = "green4")
text(220, 350, paste("Energy costs:", round(total_costs), "kcal"))
```